---
title: "FraudDetect: Financial Transaction Fraud Detection & Risk Analytics"
author: "Ludovic Njiosseu, MS Biostatistics"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: flatly
    embed-resources: true
execute:
  warning: false
  message: false
---

## Overview

This project builds a production-grade fraud detection and risk analytics pipeline on a simulated dataset of 10,000 financial transactions. It replicates core workflows used in financial services risk teams: feature engineering, class-imbalance correction, ensemble modeling, anomaly scoring, and risk segmentation.

**Key results:**

| Metric | Value |
|---|---|
| Ensemble AUC-ROC | See `outputs/model_results.csv` |
| FP Rate Reduction vs Rule-Based Baseline | See `outputs/key_metrics.csv` |
| Anomalies Flagged (top 10% isolation score) | See `outputs/risk_tier_summary.csv` |

---

## Module 1 — Data Simulation

```{r}
source("scripts/01_simulate_data.R")
```

The simulation generates realistic transaction data with known fraud-generating logic:
- **Fraud drivers**: international transactions, night-time activity, device mismatch, prior fraud history, high velocity
- **Merchant categories**: 8 categories with varying baseline risk
- **Fraud rate**: ~5–8%, reflecting real-world class imbalance

---

## Module 2 — Exploratory Analysis & Feature Engineering

```{r}
source("scripts/02_eda_features.R")
```

### Fraud by Merchant Category

```{r}
library(tidyverse)
fraud_by_cat <- read_csv("outputs/fraud_by_category.csv", show_col_types = FALSE)
fraud_by_cat %>%
  mutate(merchant_category = fct_reorder(merchant_category, fraud_rate)) %>%
  ggplot(aes(x = merchant_category, y = fraud_rate, fill = fraud_rate)) +
  geom_col() + coord_flip() +
  scale_fill_gradient(low = "#27ae60", high = "#e74c3c") +
  labs(title = "Fraud Rate by Merchant Category",
       x = NULL, y = "Fraud Rate (%)") +
  theme_minimal() + theme(legend.position = "none")
```

### Feature Correlations with Fraud

```{r}
feat_corr <- read_csv("outputs/feature_correlations.csv", show_col_types = FALSE)
feat_corr %>%
  mutate(feature = fct_reorder(feature, abs(correlation_with_fraud))) %>%
  ggplot(aes(x = feature, y = correlation_with_fraud,
             fill = correlation_with_fraud > 0)) +
  geom_col() + coord_flip() +
  scale_fill_manual(values = c("#2980b9", "#e74c3c"),
                    labels = c("Negative", "Positive")) +
  labs(title = "Feature Correlations with Fraud Label",
       x = NULL, y = "Pearson Correlation", fill = "Direction") +
  theme_minimal()
```

---

## Module 3 — Modeling

```{r}
source("scripts/03_modeling.R")
```

### Model Comparison

```{r}
model_results <- read_csv("outputs/model_results.csv", show_col_types = FALSE)
knitr::kable(model_results, caption = "AUC-ROC by Model")
```

**Approach:** Logistic regression and a bagged decision stump ensemble (RF proxy) are trained on an oversampled balanced training set (minority class oversampled to ~20%). The final prediction is a probability average of both models.

**Threshold selection:** Youden's J statistic (maximizes sensitivity + specificity − 1) is used to identify the optimal classification threshold rather than defaulting to 0.5.

### Feature Importance

```{r}
feat_imp <- read_csv("outputs/feature_importance.csv", show_col_types = FALSE)
feat_imp %>%
  mutate(feature = fct_reorder(feature, abs(coefficient)),
         direction = ifelse(coefficient > 0, "Increases Risk", "Decreases Risk")) %>%
  ggplot(aes(x = feature, y = coefficient, fill = direction)) +
  geom_col() + coord_flip() +
  scale_fill_manual(values = c("Increases Risk" = "#e74c3c",
                               "Decreases Risk" = "#2980b9")) +
  labs(title = "Logistic Regression Coefficients",
       x = NULL, y = "Log-Odds Coefficient", fill = NULL) +
  theme_minimal()
```

---

## Module 4 — Anomaly Scoring & Risk Segmentation

```{r}
source("scripts/04_anomaly_risk.R")
```

### Risk Tier Distribution

```{r}
risk_tiers <- read_csv("outputs/risk_tier_summary.csv", show_col_types = FALSE)
knitr::kable(risk_tiers, caption = "Risk Tier Summary")
```

### Top Risk Segments

```{r}
risk_segs <- read_csv("outputs/risk_segments.csv", show_col_types = FALSE)
head(risk_segs, 15) %>%
  knitr::kable(caption = "Top 15 Risk Segments by Fraud Rate")
```

**Anomaly scoring** uses an isolation-forest-inspired approach: transactions are scored by their average path length under random recursive splits. Combined with z-score thresholding on transaction amount, this yields a dual-signal anomaly flag with reduced false-positive rate versus a rule-based amount threshold alone.

---

## Reproducibility

All code is modular and documented. Run scripts in order from the project root:

```r
source("scripts/01_simulate_data.R")
source("scripts/02_eda_features.R")
source("scripts/03_modeling.R")
source("scripts/04_anomaly_risk.R")
shiny::runApp("shiny/")
```

Required packages: `tidyverse`, `pROC`, `shiny`, `bslib`, `bsicons`, `DT`

---

## Limitations & Extensions

- The dataset is simulated; real-world deployment would require calibration on live transaction data and ongoing model monitoring for distribution shift.
- The bagged stump ensemble is a pedagogical RF proxy — production deployment would use `ranger` or `xgboost`.
- Temporal features (transaction sequence, recency) are not modeled here but are critical in production fraud systems.
- No survival/time-to-fraud analysis is included; this would be a natural extension for account-level risk modeling.
